Exploring the ages of rotating stars using galactic dynamics: a novel approach to calibrating gyrochronology

### 2020-03-16 paper/conclusion.tex
adding jen's small comments and a core-envelope coupling section

\item{{\bf \sigmavb\ can be used as an age proxy for (low Galactic latitude)
    \kepler\ stars.} In the appendix of this paper we demonstrate that the
        dispersion of velocities in the direction of Galactic latitude, \vb,
        can be used as an age proxy, by showing that there is no strong
        evidence for mass-dependent heating in low-mass \kepler\ dwarfs: the
        velocity dispersions of K and M dwarfs, whose main-sequence lifetimes
        are longer than around 11 Gyrs, do not appear to increase with
        decreasing mass.
Although {\it vertical} velocity, \vz, is a quantity that has been
demonstrated to trace time-dependent orbital heating in the disc of the
Galaxy, most stars with measured rotation periods do not yet have radial
velocities, so we used velocity in the direction of Galactic latitude, \vb\,
as a proxy for \vz.
        Using stars in the GUMS simulation \citep{robin2012}, we showed that
        using \vb\ as a proxy for \vz\ introduces an additional
        velocity scatter, which increases with increasing Galactic
        latitude.
However, after removing high-latitude ($b>15^\circ$) stars from the sample, we
confirmed that using \vb\ instead of \vz\ does not introduce any
mass-dependent velocity dispersion bias into the sample.
We therefore confirmed that \vb\ velocity dispersion can be used to accurately
rank stars by age.}

>>>>

% \item{{\bf \sigmavb\ can be used as an age proxy for (low Galactic latitude)
%     \kepler\ stars.} In the appendix of this paper we demonstrate that the
%         dispersion of velocities in the direction of Galactic latitude, \vb,
%         can be used as an age proxy, by showing that there is no strong
%         evidence for mass-dependent heating in low-mass \kepler\ dwarfs: the
%         velocity dispersions of K and M dwarfs, whose main-sequence lifetimes
%         are longer than around 11 Gyrs, do not appear to increase with
%         decreasing mass.
% Although {\it vertical} velocity, \vz, is a quantity that has been
% demonstrated to trace time-dependent orbital heating in the disc of the
% Galaxy, most stars with measured rotation periods do not yet have radial
% velocities, so we used velocity in the direction of Galactic latitude, \vb\,
% as a proxy for \vz.
%         Using stars in the GUMS simulation \citep{robin2012}, we showed that
%         using \vb\ as a proxy for \vz\ introduces an additional
%         velocity scatter, which increases with increasing Galactic
%         latitude.
% However, after removing high-latitude ($b>15^\circ$) stars from the sample, we
% confirmed that using \vb\ instead of \vz\ does not introduce any
% mass-dependent velocity dispersion bias into the sample.
% We therefore confirmed that \vb\ velocity dispersion can be used to accurately
% rank stars by age.}

### 2020-03-16 paper/intro.tex
more changes
adding jen's small comments and a core-envelope coupling section

Stars with significant convective envelopes ($\lesssim$ 1.3 M$_\odot$) have
strong magnetic fields and slowly lose angular momentum via magnetic braking
\citep[\eg][]{schatzman1962, weber1967, skumanich1972, kawaler1988,
pinsonneault1989}.
Although stars are born with random rotation periods, between 1 and 10 days,
observations of young open clusters reveal that their rotation periods
converge onto a unique sequence by $\sim$500-700 million years
\citep[\eg][]{irwin2009, gallet2013}.
After this time, the rotation period of a star is thought to be determined, to
first order, by its color and age alone.
This is the principle behind gyrochronology, the method of inferring a
star’s age from its rotation period \citep[\eg][]{barnes2003, barnes2007,
barnes2010, meibom2011, meibom2015}.
However, new photometric rotation periods made available by the \kepler\
\citep{borucki2010} and \ktwo\ \citep{howell2014} missions
\citep[\eg][]{mcquillan2014, garcia2014, douglas2017, rebull2017, meibom2011,
meibom2015, curtis2019} have revealed that rotational evolution is more
complicated than previously thought.
For example, the early-to-mid M dwarfs in the $\sim$ 650 Myr Praesepe cluster
spin more slowly than the G dwarfs; in theory because lower-mass stars have
deeper convective zones which generate stronger magnetic fields and more
efficient magnetic braking.
However, in the NGC 6811 cluster which is around 1.1 Gyr \citep{janes2011},
late-K dwarfs rotate at the {\it same rate} as early-K dwarfs
\citep{curtis2019}.
In other words, convection zone depth cannot be the only variable that affects
stellar spin-down rate.
New semi-empirical models that vary the rate of angular momentum
redistribution in the interiors of stars are able to reproduce this flattened
period--color relation \citep{spada2019}.
These models suggest that mass and age-dependent angular momentum transport
between the cores and envelopes of stars has a significant impact on their
surface rotation rates.
Another example of unexpected rotational evolution is seen in old field stars
which appear to rotate more rapidly than classical gyrochronology models
predict \citep{angus2015, vansaders2016, vansaders2018, metcalfe2019}.
A mass-dependent modification to the classical \prot\ $\propto
t^{\frac{1}{2}}$ spin-down law \citep{skumanich1972} is required to reproduce
these observations.
To fit magnetic braking models to these data, a cessation of magnetic braking
is required after stars reach a Rossby number (Ro; the ratio of rotation
period to convective turnover time) of around 2 \citep{vansaders2016,
vansaders2018}.

>>>>

Stars with significant convective envelopes ($\lesssim$ 1.3 M$_\odot$) have
strong magnetic fields and slowly lose angular momentum via magnetic braking
\citep[\eg][]{schatzman1962, weber1967, kraft1967, skumanich1972, kawaler1988,
pinsonneault1989}.
Although stars are born with random rotation periods, between 1 and 10 days,
observations of young open clusters reveal that their rotation periods
converge onto a unique sequence by $\sim$500-700 million years
\citep[\eg][]{irwin2009, gallet2013}.
After this time, the rotation period of a star is thought to be determined, to
first order, by its color and age alone.
This is the principle behind gyrochronology, the method of inferring a
star’s age from its rotation period \citep[\eg][]{barnes2003, barnes2007,
barnes2010, meibom2011, meibom2015}.
However, new photometric rotation periods made available by the \kepler\
\citep{borucki2010} and \ktwo\ \citep{howell2014} missions
\citep[\eg][]{mcquillan2014, garcia2014, douglas2017, rebull2017, meibom2011,
meibom2015, curtis2019} have revealed that rotational evolution is more
complicated than previously thought.
For example, the early-to-mid M dwarfs in the $\sim$ 650 Myr Praesepe cluster
spin more slowly than the G dwarfs; in theory because lower-mass stars have
deeper convective zones which generate stronger magnetic fields and more
efficient magnetic braking.
However, in the NGC 6811 cluster which is around 1 Gyr \citep{janes2011,
sandquist2016},
late-K dwarfs rotate at the {\it same rate} as early-K dwarfs
\citep{curtis2019}.
In other words, convection zone depth cannot be the only variable that affects
stellar spin-down rate.
New semi-empirical models that vary the rate of angular momentum
redistribution in the interiors of stars are able to reproduce this flattened
period--color relation \citep{spada2019}.
These models suggest that mass and age-dependent angular momentum transport
between the cores and envelopes of stars has a significant impact on their
surface rotation rates.
Another example of unexpected rotational evolution is seen in old field stars
which appear to rotate more rapidly than classical gyrochronology models
predict \citep{angus2015, vansaders2016, vansaders2018, metcalfe2019}.
A mass-dependent modification to the classical \prot\ $\propto
t^{\frac{1}{2}}$ spin-down law \citep{skumanich1972} is required to reproduce
these observations.
To fit magnetic braking models to these data, a cessation of magnetic braking
is required after stars reach a Rossby number (Ro; the ratio of rotation
period to convective turnover time) of around 2 \citep{vansaders2016,
vansaders2018}.

### 2020-03-16 paper/method.tex
adding jen's small comments and a core-envelope coupling section

Photometric binaries and subgiants were removed from the \mct\ sample by
applying cuts to the color-magnitude diagram (CMD), shown in figure
\ref{fig:age_gradient}.
A 6th-order polynomial was fit to the main sequence and raised by 0.27 dex to
approximate the division between single stars and photometric binaries (shown
as the curved dashed line in figure \ref{fig:age_gradient}).
All stars above this line were removed from the sample.
Subgiants were also removed by eliminating stars brighter than 4th absolute
magnitude in \gaia\ G-band.

@@

The {\tt Pyia} \citep{price-whelan_2018} and {\tt astropy} \citep{astropy2013,
astropy2018} {\it Python} packages were used to calculate velocities for the
\mct\ sample.
{\tt Pyia} calculates velocity samples from the full \gaia\ uncertainty
covariance matrix via Monte Carlo sampling, thereby accounting for the
covariances between \gaia\ positions, parallaxes and proper motions.
Stars with negative parallaxes, parallax signal-to-noise ratios less than 10,
stars fainter than 16th magnitude, stars with absolute \vb\ uncertainties
greater than 1 \kms\, and stars with galactic latitudes greater than
15\degrees\ (justification provided in the appendix) were removed from the
sample.
Finally, we removed stars with rotation periods shorter than the main
population of periods, since this area of the period-\teff\ diagram is
sparsely populated.
We removed these rapid rotators by cutting out stars with gyrochronal ages
less than 0.5 Gyr \citep[based on the][gyro-model]{angus2019}, because a 0.5
Gyr gyrochrone\footnote{A gyrochrone is a gyrochronological isochrone, or a
line of constant age in period-\teff, or period-color space.} traces the
bottom edge of the main population of rotation periods.
After these cuts, around 13,000 stars out of the original $\sim$34,000 were
included in the sample.

>>>>

Photometric binaries and subgiants were removed from the \mct\ sample by
applying cuts to the color-magnitude diagram (CMD), shown in figure
\ref{fig:age_gradient}.
A 6th-order polynomial was fit to the main sequence and raised by 0.27 dex to
approximate the division between single stars and photometric binaries (shown
as the curved dashed line in figure \ref{fig:age_gradient}).
All stars above this line were removed from the sample.
Potential subgiants were also removed by eliminating stars brighter than 4th
absolute magnitude in \gaia\ G-band.
This cut also removed a number of main sequence F stars from our sample,
however these hot stars are not the focus of our gyrochronology study since
their small convective zones inhibit the generation of a strong magnetic
field.
The removal of photometric binaries and evolved/hot stars reduced the total
sample of around 34,000 stars by almost 10,000.

@@

The {\tt Pyia} \citep{price-whelan_2018} and {\tt astropy} \citep{astropy2013,
astropy2018} {\it Python} packages were used to calculate velocities for the
\mct\ sample.
{\tt Pyia} calculates velocity samples from the full \gaia\ uncertainty
covariance matrix via Monte Carlo sampling, thereby accounting for the
covariances between \gaia\ positions, parallaxes and proper motions.
Stars with negative parallaxes or parallax signal-to-noise ratios less than 10
(around 3,000 stars), stars fainter than 16th magnitude (200 stars), stars
with absolute \vb\ uncertainties greater than 1 \kms\ (1000 stars), and stars
with galactic latitudes greater than 15\degrees\ (5500 stars, justification
provided in the appendix) were removed from the sample.
Finally, we removed almost 2000 stars with rotation periods shorter than the
main population of periods, since this area of the period-\teff\ diagram is
sparsely populated.
We removed these rapid rotators by cutting out stars with gyrochronal ages
less than 0.5 Gyr \citep[based on the][gyro-model]{angus2019}, because a 0.5
Gyr gyrochrone\footnote{A gyrochrone is a gyrochronological isochrone, or a
line of constant age in period-\teff, or period-color space.} traces the
bottom edge of the main population of rotation periods.
After these cuts, around 13,000 stars out of the original $\sim$34,000 were
included in the sample.

### 2020-03-16 paper/results.tex
adding jen's small comments and a core-envelope coupling section

To explore the relationship between rotation period, effective temperature
(\teff ) and velocity dispersion, we calculated \sigmavb \footnote{\sigmavb
was calculated as 1.5$\times$ the median absolute deviation, to mitigate
sensitivity to outliers.} for groups of stars with similar rotation periods
and temperatures, and presumed similar age.
The top panel of figure \ref{fig:vplot} shows rotation period versus effective
temperature for the \mct\ sample, coloured by \sigmavb, where \sigmavb\ was
calculated for groups of stars over a grid in $\log_{10}$(period) and
temperature.
If we assume that mass dependent heating does not strongly affect this sample
and \vb\ at low galactic latitudes is an unbiased tracer of \vz, then \vb\
velocity dispersion can be interpreted as an age proxy, and stars plotted in a
similar color in figure \ref{fig:vplot} are similar ages.
In the appendix of this paper, we show that this assumption appears valid for
stars with Galactic latitude $<$ 15\degrees.
\begin{figure}
  \caption{
    Top: Rotation period vs effective temperature for stars in the \mct\
    sample, colored by the velocity dispersions of stars calculated over a
    grid in $\log_{10}$(period) and \teff.
    Black lines show gyrochrones from a gyrochronology model that projects the
    rotation-color relation of
    Praesepe to longer rotation periods over time \citep{angus2019}.
    These gyrochrones do not appear to reflect the evolution of field stars at
    long rotation periods/old ages because they do not trace lines of constant
    velocity dispersion.
    Gyrochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr (Solar age)
    in both top and bottom panels.
    Bottom: Same as top panel with rotation period vs {\it mass}
    \citep[from][]{berger2020}.
    White lines show gyrochrones from a model that includes mass and
    age-dependent angular momentum transport between the core and envelope
    \citep{spada2019}.
    Qualitatively, these gyrochrones reflect the evolution of field
    stars at long rotation periods/old ages: they trace lines of constant
    velocity dispersion by reproducing periods of `stalled' rotational
    evolution for K-dwarfs.
}
  \centering
    \includegraphics[width=1\textwidth]{main_figure}
\label{fig:vplot}
\end{figure}

@@

The results shown in figure \ref{fig:vplot} indicate that stars of spectral
type ranging from late G to late K ($\sim$5500-3500 K) follow a braking law
that changes over time.
In particular, the relationship between rotation period and effective
temperature appears to flatten out and eventually invert.
These results provide further evidence for `stalled' rotational evolution of K
dwarfs, like that observed in open clusters \citep{curtis2019} and reproduced
by models that vary angular momentum transport between stellar core and
envelope with time and mass \citep{spada2019}.
The velocity dispersions of stars in the \mct\ sample provide the following
picture of rotational evolution.
At young ages \citep[younger than around 1 Gyr but still old enough to be on
the main sequence and have transitioned from the `I' sequence to the `C'
sequence ][]{barnes2003}, stellar rotation period {\it decreases} with {\it
increasing} mass.
This is likely because lower-mass stars with deeper convection zones have
stronger magnetic fields, larger Alfv\'en radii and therefore experience
greater angular momentum loss rate \citep[\eg][]{schatzman1962, parker1970,
kawaler1988, charbonneau2010}.
According to the \citet{spada2019} model, there is minimal transportation of
angular momentum from the surface to the core of the star at these young ages,
so the surface slows down but the core keeps spinning rapidly.
This aligns with the current assumptions about stellar spin-down, dynamo
theory, and the gyrochronology paradigm that has been in place for decades
\citep[\eg][]{skumanich1972, noyes1984, kawaler1988, barnes2003, angus2019}.
According to the data presented in figure \ref{fig:vplot} however, at
intermediate ages, the rotation periods of K dwarfs appear {\it constant} with
mass, and at late ages rotation period {\it increases} with {\it increasing}
mass.
The interpretation of this, according to the \citet{spada2019} model, is that
lower-mass stars are still braking more efficiently at these intermediate and
old ages but their cores are more tightly coupled to their envelopes, allowing
angular momentum transport between the two layers.
Angular momentum resurfaces and prevents the stellar envelopes from
spinning-down rapidly, and this effect is strongest for late K-dwarfs with
effective temperatures of $\sim$4000-4500K and masses $\sim$0.5-0.7 M$_\odot$.

It has been demonstrated that lower-mass stars remain magnetically active
longer than more massive stars, \citep[\eg][]{west2008, kiman2019}.
If the detectability of a rotation period is considered to be a magnetic
activity proxy, then our results provide further evidence for a mass-dependent
activity lifetime.
Figure \ref{fig:vplot} shows that the groups of stars with the largest
velocity dispersions are cooler than 4500 K.
This implies that the oldest stars with detectable rotation periods, are
cooler than 4500 K, \ie\ these-low mass stars stay active longer than more
massive stars.
To investigate this idea further, we compared the velocity dispersions of
stars with measured rotation periods, to the velocity dispersions of the
overall \kepler\ sample.
We calculated the velocity dispersions for all stars in the \kepler\ field,
after removing visual binaries, subgiants, stars fainter than 16th magnitude,
and high Galactic latitude stars, following the method described in section
\ref{sec:method}.
We then compared these velocity dispersions to the velocity dispersions of
stars in the \mct\ sample.
If the rotation periods of G stars are only detectable when they are young,
G stars with measured periods should have smaller velocity dispersions than
the overall \kepler\ field.
We calculated the ratio of \sigmavb\ for the entire \Kepler\ sample to
\sigmavb\ for stars with rotation periods published in \mct, as a function of
\teff\ (see figure \ref{fig:compare}).
A larger ratio means the rotating star sample is {\it younger}, on average,
than the overall \Kepler\ population, and a ratio of 1 means that the rotating
stars have the {\it same} age distribution as the overall Kepler sample.
Figure \ref{fig:compare} shows that this ratio is largest for G stars and
approaches unity for K and early M dwarfs.
This indicates that the G stars with detectable rotation periods are, on
average, {\it younger} than the total population of G stars in the Kepler
field.
On the other hand, the late K and early M dwarfs with detectable rotation
periods have a similar age distribution to the overall Kepler population which
suggests that the oldest K and M dwarfs are represented in the \mct\ sample.
This result bolsters the evidence that M dwarf rotation periods are measurable
at older ages than G dwarf rotation periods.
G stars become magnetically inactive and have fewer active surface regions
{\it at a younger age than M dwarfs}.

>>>>

To explore the relationship between rotation period, effective temperature
(\teff ) and velocity dispersion, we calculated \sigmavb \footnote{\sigmavb
was calculated as 1.5$\times$ the median absolute deviation, to mitigate
sensitivity to outliers.} for groups of stars with similar rotation periods
and temperatures, and presumed similar age.
The top panel of figure \ref{fig:vplot} shows rotation period versus effective
temperature for the \mct\ sample, coloured by \sigmavb, where \sigmavb\ was
calculated for groups of stars over a grid in $\log_{10}$(period) and
temperature.
If we assume that mass dependent heating does not strongly affect this sample
and \vb\ at low galactic latitudes is an unbiased tracer of \vz, then \vb\
velocity dispersion can be interpreted as an age proxy, and stars plotted in a
similar color in figure \ref{fig:vplot} are similar ages.
In the appendix of this paper, we show that this assumption appears valid for
stars with Galactic latitude $<$ 15\degrees.
\begin{figure}
  \caption{
    Top: Rotation period vs effective temperature for stars in the \mct\
    sample, colored by the velocity dispersions of stars calculated over a
    grid in $\log_{10}$(period) and \teff\ (this grid causes the quantized
    appearance).
    Black lines show gyrochrones from a gyrochronology model that projects the
    rotation-color relation of
    Praesepe to longer rotation periods over time \citep{angus2019}.
    These gyrochrones do not appear to reflect the evolution of field stars at
    long rotation periods/old ages because they do not trace lines of constant
    velocity dispersion.
    Gyrochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr (Solar age)
    in both top and bottom panels.
    Bottom: Same as top panel with rotation period vs {\it mass}
    \citep[from][]{berger2020}.
    White lines show gyrochrones from a model that includes mass and
    age-dependent angular momentum transport between the core and envelope
    \citep{spada2019}.
    Qualitatively, these gyrochrones reflect the evolution of field
    stars at long rotation periods/old ages: they trace lines of constant
    velocity dispersion by reproducing periods of `stalled' surface rotational
    evolution for K-dwarfs.
}
  \centering
    \includegraphics[width=1\textwidth]{main_figure}
\label{fig:vplot}
\end{figure}

@@

The results shown in figure \ref{fig:vplot} indicate that stars of spectral
type ranging from late G to late K ($\sim$5500-3500 K) follow a braking law
that changes over time.
In particular, the relationship between rotation period and effective
temperature appears to flatten out and eventually invert.
These results provide further evidence for `stalled' surface rotational
evolution of K dwarfs, like that observed in open clusters \citep{curtis2019}
and reproduced by models that vary angular momentum transport between stellar
core and envelope with time and mass \citep{spada2019}.
The velocity dispersions of stars in the \mct\ sample provide the following
picture of rotational evolution.
At young ages \citep[younger than around 1 Gyr but still old enough to be on
the main sequence and have transitioned from the `I' sequence to the `C'
sequence ][]{barnes2003}, stellar rotation period {\it decreases} with {\it
increasing} mass.
This is likely because lower-mass stars with deeper convection zones have
stronger magnetic fields, larger Alfv\'en radii and therefore experience
greater angular momentum loss rate \citep[\eg][]{schatzman1962, parker1970,
kawaler1988, charbonneau2010}.
According to the \citet{spada2019} model, there is minimal transportation of
angular momentum from the surface to the core of the star at these young ages,
so the surface slows down but the core keeps spinning rapidly.
This aligns with the current assumptions about stellar spin-down, dynamo
theory, and the gyrochronology paradigm that has been in place for decades
\citep[\eg][]{skumanich1972, noyes1984, kawaler1988, barnes2003, angus2019}.
According to the data presented in figure \ref{fig:vplot} however, at
intermediate ages, the rotation periods of K dwarfs appear {\it constant} with
mass, and at late ages rotation period {\it increases} with {\it increasing}
mass.
The interpretation of this, according to the \citet{spada2019} model, is that
lower-mass stars are still braking more efficiently at these intermediate and
old ages but their cores are more tightly coupled to their envelopes, allowing
angular momentum transport between the two layers.
Angular momentum resurfaces and prevents the stellar envelopes from
spinning-down rapidly, and this effect is strongest for late K-dwarfs with
effective temperatures of $\sim$4000-4500K and masses $\sim$0.5-0.7 M$_\odot$.

A period of core-envelope decoupling in the evolution of cool dwarfs has been
explored in theoretical models for decades \citep[\eg][]{endal1981,
macgregor1991, denissenkov2010, gallet2013}.
In such models, the angular momenta of the radiative core and convective
envelope are permitted to evolve separately once a radiative core develops on
the pre-main sequence.
A core-envelope decoupling phase in the early angular momentum history of
stars is required in order to reproduce observations of young clusters and
star forming regions \citep[\eg][]{irwin2007, bouvier2008, denissenkov2010,
spada2011, reiners2012} and has become an established element of theoretical
gyrochronology models.
The \citet{lanzafame2015} gyrochronology model quantified the mass-dependence
of the core-envelope coupling timescale and this model was fit to open-cluster
observations in \citet{spada2019}, including new rotation period measurements
for stars down to M$\sim$0.6M$_\odot$ in the NGC 6811 cluster
\citep{curtis2019}.
A similar mass-dependent core-envelope coupling timescale was also found to
explain the observed lithium depletion trends in open clusters by an
independent study \citep{somers2016}.
Although variable angular momentum transport between the surfaces and cores of
stars has been an essential ingredient of stellar evolution models for
decades, the transport mechanism is still unknown.
Among the proposed mechanisms are magneto-hydrodynamical waves resulting from
various magnetic field geometries, and gravity waves \citep[see,
\eg][]{charbonneau1993, ruediger1996, spruit2002, talon2003, spada2010,
brun2011, oglethorpe2013}.

Figure \ref{fig:vplot} reveals another phenomenon of magnetic evolution:
K and M dwarfs remain magnetically active for longer than G dwarfs.
The mass dependence of magnetic activity lifetimes has been demonstrated
previously \citep[\eg][]{west2008, newton2017, kiman2019}.
If the detectability of a rotation period is considered to be a magnetic
activity proxy, then our results provide further evidence for a mass-dependent
activity lifetime.
Figure \ref{fig:vplot} shows that the groups of stars with the largest
velocity dispersions are cooler than 4500 K.
This implies that the oldest stars with detectable rotation periods, are
cooler than 4500 K, \ie\ these-low mass stars stay active longer than more
massive stars.
To investigate this idea further, we compared the velocity dispersions of
stars with measured rotation periods, to the velocity dispersions of the
overall \kepler\ sample.
We calculated the velocity dispersions for all stars in the \kepler\ field,
after removing visual binaries, subgiants, stars fainter than 16th magnitude,
and high Galactic latitude stars, following the method described in section
\ref{sec:method}.
We then compared these velocity dispersions to the velocity dispersions of
stars in the \mct\ sample.
If the rotation periods of G stars are only detectable when they are young,
G stars with measured periods should have smaller velocity dispersions than
the overall \kepler\ field.
We calculated the ratio of \sigmavb\ for the entire \Kepler\ sample to
\sigmavb\ for stars with rotation periods published in \mct, as a function of
\teff\ (see figure \ref{fig:compare}).
A larger ratio means the rotating star sample is {\it younger}, on average,
than the overall \Kepler\ population, and a ratio of 1 means that the rotating
stars have the {\it same} age distribution as the overall Kepler sample.
Figure \ref{fig:compare} shows that this ratio is largest for G stars and
approaches unity for K and early M dwarfs.
This indicates that the G stars with detectable rotation periods are, on
average, {\it younger} than the total population of G stars in the Kepler
field.
On the other hand, the late K and early M dwarfs with detectable rotation
periods have a similar age distribution to the overall Kepler population which
suggests that the oldest K and M dwarfs are represented in the \mct\ sample.
This result bolsters the evidence that M dwarf rotation periods are measurable
at older ages than G dwarf rotation periods.
G stars become magnetically inactive and have fewer active surface regions
{\it at a younger age than M dwarfs}.
